Incremental tactics

### 2012-05-30 doc/papers/LFMTP2012/paper.tex
paper

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to an already-computed value;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

\section{Background}
%% It’s an unsolved problem (1 page)

\paragraph{Motivational example}

@@

@@

\subsection{The representation language}

@@

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax. This framework has been used to issue certificates for a
number of practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{0}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$\app {\cst{lam}} \lam x
\fbox{$\app {\cst{lam}} \lam y
\fbox{\app {\cst{app}} \app {\var y}
}$}
{(\fbox{\app
  {\cst{app}}\app{\var x}{\cst 0}})
}
$}$$
where all boxed subterms are referable slices.

\subsection{Computational calculus}
\subsection{Slicing results of computation}

>>>>

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to already-computed values;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

\section{Use case}
%% It’s an unsolved problem (1 page)

\subsection{A backend for incremental type checking}
\label{sec:use-incremental}

@@

\subsection{A language to write incremental type-checkers}

A few days earlier, the teacher is preparing the tool for the students
to train composing System \sysname{T} programs. He (or she) provides
the abstract syntax, the typing rules and the typing untrusted
algorithm in order to build the incremental type checker. How should
he write the algorithm? We know that we want to return a derivation,
and take advantage of a higher-order representation of the syntax, so
it shouldn't be necessary to thread any explicit environment. The
definition should then start with:

\begin{mathleft}
  \finfer{}\ :\ \prd M {\cst{tm}} \sig A {\cst{tp}} (\vdash M : A) =
\end{mathleft}

\noindent
It takes a term $M$ and producing the pair of a type $A$ and a
derivation that $M$ has type $A$. Its code will be untyped, but
the teacher provides this type so that the checker can verify that
each call is fed and returns values of the right type: it is its
\emph{contract} so to say. He first decomposes term $M$ and treats
the easy cases of $\cst{o}$, $\cst{s}$ and $\cst{app}$:

\begin{mathleft}
  \lamd M \match{M} \\
  \quad\case{\z} \pair {\cst{nat}}{\infer{ }{\vdash \cst o : \cst{nat}}} \\
  \quad\case{\s M}
  \letd {\pair {\cst{nat}} \md} {\finfer{M}}
  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
  \quad\case{\app{M}N}
  \letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}}
\end{mathleft}

\noindent
We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types; the
$\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; patterns don't have to be exhaustive since we are in
an untrusted setting; for conciseness, we use case failure for
signaling typing error; we use an auxiliary function $\!\!\feq{}{}\!\!
: \cst{tp}\to\cst{tp}\to\cst{bool}$ for syntactic comparison of types,
defined the same way.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$, otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
call, arriving to variable $\var x$ would fail since we would have no
way of telling its type anymore. Thus we substitute all occurrences of
$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
substitution to be well-typed, we now need to coerce this output to a
valid input (a term). We introduce an operator which is in some way
the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
type and a derivation to the term it types.

\begin{mathleft}
  \quad\case{\tlam x A M} \\
  \quad\quad
  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
\end{mathleft}

At this point, we get back $B$ and $\md$ from the recursive call. We
know that $B$ can't depend on $\md_x$ (because a type is always
closed), but $\md$ can: it is not \emph{per se} a hypothetical
derivation but an \emph{open} derivation where a special judgement
named $\md_x$ can occur. To return the complete derivation of
$\vdash\tlam A x M : A \to B$, we must then close, or ``hypothesize''
$\md$ by $\md_x$, and then apply rule \rulename{Lam}:

\begin{mathleft}
  \quad\quad
  \pair {A\to B}
  {\infer{
\mbox{$      \begin{array}{c}
        {[\md_{\var x}]} \\
        {\md}
      \end{array}
    $}    }{
    \vdash\lam x M : A\to B
  }}
\end{mathleft}

The case of $\cst{rec}$ is similar: we first compute recursively the
derivations $\md_M$ and $\md_N$ which should be well-typed in the
current environment, and then $\md_P$ in the enlarged environment,
with variables $\var x$ and $\var y$ in $P$ substituted with the
(future) result of $\finfer{}$.

\begin{mathleft}
  \quad\case{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {\cst{nat}} {\md_M}} {\finfer{M}} \\
  \quad\quad
  \letd {\pair {A_1} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_2} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_1)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_1}{\md_y}}}}}} \\
  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}} \\
  \quad\quad
  \pair {A} {
    \infer{
      \md_M \and
      \md_N \and
      \mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}
    }{
      \vdash \recb M N x y P
    }
  }
\end{mathleft}

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. We only need to make sure that in this case,
the derivation will be returned:
$$
\finfer{\fget{\pair A \md}} = \md
$$
Now $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \\\quad \lamd M \lamd A \lamd \md M
\end{mathleft}

\noindent
where argument $M$ is implicit.

We exploit here an idea similar to run-time
\emph{contracts}: %TODO cite
there is no static guarantee that computations won't return ill-typed
values, but arguments and results are checked at run-time at the
boundaries of defined functions. This allows to omit formal
justifications, like the fact that $B$ can't depend on $\md_x$ in the
$\lambda$ case, but of course incurs on the performance. This is
compensated by the slicing mechanism described below in
\ref{sec:repr}.

Introducing the ``inverse'' of the derivation inference function has
two purpose: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones.

The difficulty with this technique, particularly in a dependent
setting, is to ensure that term $\{\var x:A\}$ reduces to $\var x$ so
that they are considered equal, \ie\ we can strip off the annotations,
while not forgetting them too soon: otherwise we would arrive to the
unhandled case $\finfer{\var x}$. This requires evaluating the
type-checker with a carefully designed strategy.

@@

\subsection{The representation language}
\label{sec:repr}

@@

The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is a well-accepted solution to this question. It is a typed
lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.

For instance, the rule for System \sysname{T}'s recursor can be
encoded by the following constant:

\begin{mathleft}
  \cst{is\_rec}\ :\
  \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
  \\\quad
  \prd A {\cst{tp}}
  \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
  \left(
  \prd x {\cst{tm}} \prd y {\cst{tm}}
  \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
  \app{\cst{is}}\app {(\app P \app x y)} A
  \right)\to\\\quad
  \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
    P\app x y))} A
\end{mathleft}

This framework has been used to issue certificates for a
number of practical applications like proof-carrying code
\cite{necula1997proof}, authentication framework \cite{appel1999proof}
\ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm and instantiate its free variables. For this, we define the
the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
metavariables $\meta X$, $\meta Y$\ldots\ as in
\cite{nanevski2008contextual}. Metavariables stand for well-typed
terms defined in a repository, and are instantiated wrt. their free
variables. A repository also contains a distinguished metavariable
corresponding to the head of the term stored in it. For instance, the
repository
$$
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing of the term
$$
\fbox{$\app {\cst{lam}} \lam x
\fbox{$\app {\cst{lam}} \lam y
\fbox{\app {\cst{app}} \app {\var y}
}$}
{(\fbox{\app
  {\cst{app}}\app{\var x}{\cst o}})
}
$}$$
where all boxed subterms are referable slices.

\subsection{The computational calculus}
